---
title: "RNASeq"
author:
  - "Victoria Ríos"
  - "Antonio Martínez"
date: "`r format(Sys.Date(),'%B %d, %Y')`"
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Import RNASeq data frame
```{r}
data <- read.csv("TNBC__maseqv2.csv",stringsAsFactors=FALSE,header=FALSE,row.names=1)
colnames(data) <- data[2,] #Define column names
data <- data[-c(1:3),] #Remove trash
df.RNAseq <- data.frame(lapply(data,as.numeric),row.names=rownames(data)) #Make all columns numeric vectors

class(df.RNAseq) # Verify its class
str(df.RNAseq,list.len=10) # Check its structure
```

Create the transposed RNASeq data frame  (columns <-> rows)
```{r}
library(data.table)
df.RNAseq.t <- transpose(df.RNAseq) # transposed data frame
colnames(df.RNAseq.t)<-rownames(df.RNAseq)
rownames(df.RNAseq.t)<-colnames(df.RNAseq)

class(df.RNAseq.t) # Verify its class
str(df.RNAseq.t, list.len=10) # Check its structure
```

Importamos el data frame de la información con los id's de los pacientes que son Triple Negativo
```{r}
df.TNBC.id <- read.csv("TNBC_merged_id.csv") # Data frame with TNBC patients' IDs
df.id <- gsub("-",".",df.TNBC.id$patient.bcr_patient_barcode_upper)
class(df.id) # Verify its class
str(df.id) # Check its structure
```
Filter the values of the transposed data frame (df.RNASeq.t) to keep only the rows of information that match the IDs of the TNBC patients (df.TNBC.id)
```{r}
library(dplyr)

match.df <- paste("^(", paste0(df.id, collapse = "|"), ")", sep = "")

tnbc.df <- filter(df.RNAseq.t, Reduce("|", lapply(rownames(df.RNAseq.t), grepl, pattern = match.df)))
```

Check the structure of the resulting filtered data frame
```{r}
str(tnbc.df, list.len = 10) # Check its structure
```

