---
title: "Code_template_normalized"
author: "Victoria Rios"
date: "12/4/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Análisis de datos de BRCA TN TCGA

Primero definimos la opción para no se lean los strings como factores.

```{r}
options(stringsAsFactors = F)
```

Primero leemos los datos clínicos.
```{r}
brca_clin <- read.delim(file = "BRCA.clin.merged.txt", sep = "\t", header = F,
row.names = 1, as.is = T)
```

Obtenemos los datos para los 3 receptores en los renglones 24, 29 y 1160.
```{r}
print(rownames(brca_clin)[c(24,29,1160)])
```

Índices de muestras triple negativo (TN)
```{r}
index_TN <- which(as.character(brca_clin[24,]) == "negative" &
as.character(brca_clin[29,]) == "negative" &
as.character(brca_clin[1160,]) == "negative")
```

Filtramos para tener datos sólo de los TN y obtenemos las claves de las muestras

```{r}
brca_clin <- brca_clin[,index_TN]
tn_samples <- toupper(as.character(brca_clin[21,]))
```

Leer datos de RNA-seq
```{r}
f <- "BRCA.rnaseqv2__illuminahiseq_rnaseqv2__unc_edu__Level_3__RSEM_genes_normalized__data.data.txt"
brca_rnaseq <- read.delim(file = f, sep="\t", as.is=T, row.names = 1)
brca_rnaseq <- data.matrix(brca_rnaseq[-1,])
```

Extraer gene symbols
```{r}
gene_symbols <- as.character(sapply(rownames(brca_rnaseq), function(x)
unlist(strsplit(x = x, split = "\\|"))[1]))
```

Eliminar renglones con gene symbol igual a “?”
```{r}
brca_rnaseq <- brca_rnaseq[which(gene_symbols != "?"),]
gene_symbols <- gene_symbols[which(gene_symbols != "?")]
```

Ver el tipo de muestra de acuerdo al TCGA barcode y quitar los normales (tipo “11”) https://docs.gdc.
cancer.gov/Encyclopedia/pages/TCGA_Barcode/
```{r}
sample_type <- substr(x = colnames(brca_rnaseq), start = 14, stop = 15)
brca_rnaseq <- brca_rnaseq[,-which(sample_type == "11")]
sample_type <- sample_type[-which(sample_type == "11")]
```

Hacer match entre las muestras TN y los de RNAseq
```{r}
tn_match <- match(gsub(pattern = "-", replacement = "\\.", x = tn_samples),
substr(x = colnames(brca_rnaseq), start = 1, stop = 12))
tn_match_na <- which(is.na(tn_match))
brca_clin <- brca_clin[,-tn_match_na]
tn_samples <- tn_samples[-tn_match_na]
tn_match <- tn_match[-tn_match_na]
brca_rnaseq <- brca_rnaseq[,tn_match]

brca_rnaseq_t <- t(brca_rnaseq) # Transpose of Rnaseq 

```

# Normalization of the dataset
```{r}
brca_rnaseq_n <- log2(brca_rnaseq_t + 1)
```

# Clustering PCA results (normalized)

## PCA
```{r}
# Principal components
x <- prcomp(brca_rnaseq_n, scale = FALSE)
```

```{r}
library(factoextra)

#Visualize eigenvalues (scree plot). Show the percentage of variances explained by each principal component.
fviz_eig(x)
```
## Graph of individuals. 
Individuals with a similar profile are grouped together.
```{r}
fviz_pca_ind(x,
             col.ind = "cos2", # Color by the quality of representation
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = T,     # Avoid text overlapping,
             max.overlaps = Inf
             )
```
## Re-labeling the patients IDs
```{r}
# We check that the patient ID doesn't repeat
temp <- table(x[["x"]])

isTRUE(any(temp > 1)) # if the result is FALSE, they appear only once
```
We save the original IDs in another list inside of x
```{r}
orig_IDs <- matrix(row.names(x[["x"]]), ncol = 1) # we store the original IDs
short_IDs <- as.character(matrix(data = 1:nrow(x[["x"]]), ncol = 1)) # we store the short IDs

IDs <- cbind(orig_IDs, short_IDs) # we bind them in the same matrix


x[["IDs"]] <- IDs # We append the new IDs matrix to the x list
```

We proceed to assign a unique number to each patient
```{r}
row.names(x[["x"]]) <- short_IDs

head(x[["x"]])
```

We plot again to visualize the short IDs of each patient
```{r}
fviz_pca_ind(x,
             col.ind = "cos2", # Color by the quality of representation
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = T,     # Avoid text overlapping,
             max.overlaps = Inf
             )
```

## Graph of variables. 
Positive correlated variables point to the same side of the plot. Negative correlated variables point to opposite sides of the graph.
```{r}
fviz_pca_var(x,
             col.var = "contrib", # Color by contributions to the PC
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE     # Avoid text overlapping
             )
```

## Biplot of individuals and variables
```{r}
fviz_pca_biplot(x, repel = TRUE,
                col.var = "#2E9FDF", # Variables color
                col.ind = "#696969"  # Individuals color
                )
```
 

## K-means with PCA results --------------------------------------------------------------------

```{r}
# Coordinates of individuals
ind.coord <- as.data.frame(get_pca_ind(x)$coord)

head(ind.coord)
```
# Let's find the optimal K value for our data

# 1. Elbow method
```{r}
library(purrr)

# function to compute total within-cluster sum of square 
wss <- function(k) {
  kmeans(ind.coord, k, nstart = 10 )$tot.withinss
}

# Compute and plot wss for k = 1 to k = 15
k.values <- 1:15

# extract wss for 2-15 clusters
wss_values <- map_dbl(k.values, wss)

plot(k.values, wss_values,
       type="b", pch = 19, frame = FALSE, 
       xlab="Number of clusters K",
       ylab="Total within-clusters sum of squares")
```
## Elbow method (Automatic)
```{r}
fviz_nbclust(ind.coord, kmeans, method = "wss") +
  labs(subtitle = "Elbow method")
```

# 2. Average Silhouette Method (Manual)
 A high average silhouette width indicates a good clustering. The average silhouette method computes the average silhouette of observations for different values of k. The optimal number of clusters k is the one that maximizes the average silhouette over a range of possible values for k.
```{r}
library(cluster)

# function to compute average silhouette for k clusters
avg_sil <- function(k) {
  km.res <- kmeans(ind.coord, centers = k, nstart = 25)
  ss <- silhouette(km.res$cluster, dist(ind.coord))
  mean(ss[, 3])
}

# Compute and plot wss for k = 2 to k = 15
k.values <- 2:15

# extract avg silhouette for 2-15 clusters
avg_sil_values <- map_dbl(k.values, avg_sil)

plot(k.values, avg_sil_values,
       type = "b", pch = 19, frame = FALSE, 
       xlab = "Number of clusters K",
       ylab = "Average Silhouettes")
```
## Average silhouette (Automatic)
```{r}
fviz_nbclust(ind.coord, kmeans, method = "silhouette")+
  labs(subtitle = "Silhouette method")
```
#3. Gap Statistic Method
```{r}
# compute gap statistic
gap_stat <- clusGap(ind.coord, FUN = kmeans, nstart = 25,
                    K.max = 10, B = 50)
# Print the result
print(gap_stat, method = "firstmax")
```
```{r}
# We visualize the K value suggestion for this method
fviz_gap_stat(gap_stat) +
  labs(subtitle = "Gap statistic method")
```

# Results
With most of these approaches suggesting 2 as the number of optimal clusters, we can perform the final analysis and extract the results using 2 clusters.

```{r}
kc <- kmeans(ind.coord, centers = 2, nstart = 25)

plot(ind.coord[,1:2],col=factor(kc$cluster))
```

```{r}
# Add clusters obtained using the K-means algorithm
ind.coord$cluster <- factor(kc$cluster)
```

```{r}
# Percentage of variance explained by dimensions
eigenvalue <- round(get_eigenvalue(x), 1)
variance.percent <- eigenvalue$variance.percent

head(eigenvalue)
```


```{r}
library(ggpubr)

ggscatter(
  ind.coord, x = "Dim.1", y = "Dim.2", 
  color = "cluster", palette = "npg", ellipse = TRUE, ellipse.type = "convex", size = 1.5,  legend = "right", ggtheme = theme_bw(),
  xlab = paste0("Dim 1 (", variance.percent[1], "% )" ),
  ylab = paste0("Dim 2 (", variance.percent[2], "% )" )
) +
  stat_mean(aes(color = cluster), size = 4)
```

# Comparing the results with different K values

```{r}
ind.coord_k2 <- ind.coord
ind.coord_k3 <- ind.coord
ind.coord_k4 <- ind.coord
ind.coord_k5 <- ind.coord

k2 <- kmeans(ind.coord_k2, centers = 2, nstart = 25)
k3 <- kmeans(ind.coord_k3, centers = 3, nstart = 25)
k4 <- kmeans(ind.coord_k4, centers = 4, nstart = 25)
k5 <- kmeans(ind.coord_k5, centers = 5, nstart = 25)

# Add clusters obtained using the K-means algorithm
ind.coord_k2$cluster <- factor(k2$cluster)
ind.coord_k3$cluster <- factor(k3$cluster)
ind.coord_k4$cluster <- factor(k4$cluster)
ind.coord_k5$cluster <- factor(k5$cluster)



# plots to compare
plot(ind.coord[,1:2],col=factor(k2$cluster))
plot(ind.coord[,1:2],col=factor(k3$cluster))
plot(ind.coord[,1:2],col=factor(k4$cluster))
plot(ind.coord[,1:2],col=factor(k5$cluster))

```
```{r}
ggscatter(
  ind.coord_k2, x = "Dim.1", y = "Dim.2", 
  color = "cluster", palette = "npg", ellipse = TRUE, ellipse.type = "convex", size = 1.5,  legend = "right", ggtheme = theme_bw(),
  xlab = paste0("Dim 1 (", variance.percent[1], "% )" ),
  ylab = paste0("Dim 2 (", variance.percent[2], "% )" )
) +
  stat_mean(aes(color = cluster), size = 4)
```

```{r}
ggscatter(
  ind.coord_k3, x = "Dim.1", y = "Dim.2", 
  color = "cluster", palette = "npg", ellipse = TRUE, ellipse.type = "convex", size = 1.5,  legend = "right", ggtheme = theme_bw(),
  xlab = paste0("Dim 1 (", variance.percent[1], "% )" ),
  ylab = paste0("Dim 2 (", variance.percent[2], "% )" )
) +
  stat_mean(aes(color = cluster), size = 4)
```

```{r}
ggscatter(
  ind.coord_k4, x = "Dim.1", y = "Dim.2", 
  color = "cluster", palette = "npg", ellipse = TRUE, ellipse.type = "convex", size = 1.5,  legend = "right", ggtheme = theme_bw(),
  xlab = paste0("Dim 1 (", variance.percent[1], "% )" ),
  ylab = paste0("Dim 2 (", variance.percent[2], "% )" )
) +
  stat_mean(aes(color = cluster), size = 4)
```

```{r}
ggscatter(
  ind.coord_k5, x = "Dim.1", y = "Dim.2", 
  color = "cluster", palette = "npg", ellipse = TRUE, ellipse.type = "convex", size = 1.5,  legend = "right", ggtheme = theme_bw(),
  xlab = paste0("Dim 1 (", variance.percent[1], "% )" ),
  ylab = paste0("Dim 2 (", variance.percent[2], "% )" )
) +
  stat_mean(aes(color = cluster), size = 4)
```



