---
title: "Characterization of expresion data"
author: "Victoria Rios"
date: "4/16/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Importing the libraries
options(stringsAsFactors=F)
library(limma)
library(sva)
library(preprocessCore)
```

## Loading the TNBC datasets

```{r}
# Importing the datasets
setwd("C:/Users/victo/OneDrive/Documentos/TNBC_research/Database exploring/DB compressed files/")

TCGA_db <- read.csv("TCGA/TCGA_TNBC_exprs.csv", row.names = "row.names")
#TCGA_db <- as.data.frame(t(TCGA_db))
#Ref1_db <- read.csv("GSE87049/TCGA_TNBC_exprs.csv") #NA (lacking of TNBC data)
Ref2_db <- read.csv("Ref 2/Ref2_untar/brca_metabric/Ref2_TNBC_exprs.csv", row.names = "row.names")
#Ref3_db <- read.csv("Ref 3/Ref3_untar/Ref3_TNBC_exprs.csv", row.names = "row.names")
Ref4_db <- read.csv("Ref 4/Ref4_untar/brca_mbcproject_wagle_2017/Ref4_TNBC_exprs.csv", row.names = "row.names")
RefGSE20271_db <- read.csv("GSE20271/GSE20271_TNBC_exprs.csv", row.names = "row.names")
RefGSE575678_db <- read.csv("GSE575678/GSE75678_TNBC_exprs.csv", row.names = "row.names")
```

## Loading clinical data
```{r}
TCGA_clin <- read.csv("C:\\Users\\victo\\OneDrive\\Documentos\\TNBC_research\\Database exploring\\DB compressed files\\TCGA\\TCGA_TNBC_clinical_complete.csv", row.names = "row.names")
TCGA_clin$source_db <- rep(x = "TCGA_RNAseq", times=nrow(TCGA_clin))

#Ref1_db <- read.csv("GSE87049/TCGA_TNBC_exprs.csv") #NA (lacking of TNBC data)

Ref2_clin <- read.csv("C:\\Users\\victo\\OneDrive\\Documentos\\TNBC_research\\Database exploring\\DB compressed files\\Ref 2\\Ref2_untar\\brca_metabric\\Ref2_TNBC_clinical_complete.csv", row.names = "row.names")
Ref2_clin$source_db <- rep(x = "Ref2_RNAseq", times=nrow(Ref2_clin))
Ref2_clin <- Ref2_clin[intersect(colnames(Ref2_db), rownames(Ref2_clin)),]

#Ref3_clin <- read.csv("Ref 3/Ref3_untar/Ref3_TNBC_clinical.csv", row.names = "row.names")

Ref4_clin <- read.csv("C:\\Users\\victo\\OneDrive\\Documentos\\TNBC_research\\Database exploring\\DB compressed files\\Ref 4\\Ref4_untar\\brca_mbcproject_wagle_2017\\Ref4_TNBC_clinical_complete.csv", row.names = "row.names")
Ref4_clin$source_db <- rep(x = "Ref4_RNAseq", times=nrow(Ref4_clin))
Ref4_clin <- Ref4_clin[intersect(colnames(Ref4_db), rownames(Ref4_clin)),]

RefGSE20271_clin <- read.csv("C:\\Users\\victo\\OneDrive\\Documentos\\TNBC_research\\Database exploring\\DB compressed files\\GSE20271\\GSE20271_TNBC_clinical.csv", row.names = "row.names")
RefGSE20271_clin$source_db <- rep(x = "RefGSE20271_Array", times=nrow(RefGSE20271_clin))

RefGSE575678_clin <- read.csv("C:\\Users\\victo\\OneDrive\\Documentos\\TNBC_research\\Database exploring\\DB compressed files\\GSE575678\\GSE75678_TNBC_clinical_complete.csv", row.names = "row.names")
RefGSE575678_clin$source_db <- rep(x = "RefGSE575678_Array", times=nrow(RefGSE575678_clin))
```

## PCA

```{r}
library(ggrepel)
library(factoextra)

## MAKE SURE TO TRANSPOSE THE EXPRESSION MATRIX
age <- as.factor(TCGA_clin$age)
race <- as.factor(TCGA_clin$race)
#treatment <- as.factor(TCGA_clin$treatment)
#methastasis <- as.factor(sampleInfo$methastasis)

pca <- prcomp(t(TCGA_db))

## Join the PCs to the sample information
fviz_pca_ind(pca, repel=T, col.ind=age, label="none")
#fviz_pca_ind(pca, repel=T, col.ind=race, label="none")
#fviz_pca_ind(pca, repel=T, col.ind=treatment, label="none")
#fviz_pca_ind(pca, repel=T, col.ind=methastasis, label="none")
```
## AGE
```{r}
head(TCGA_clin)
```
### TCGA
```{r}
boxplot(TCGA_clin$age)
```

```{r}
boxplot(RefGSE575678_clin$age)
```

## HISTOLOGICAL TYPE

### TCGA
```{r}
table(TCGA_clin$histological_type)
```

### Ref2
```{r}
table(Ref2_clin$histological_type)
```

### Ref4
```{r}
table(Ref4_clin$histological_type)
```

### RefGSE20271
```{r}
table(RefGSE20271_clin$histological_type)
```

## HISTOLOGICAL GRADE

### TCGA
```{r}
table(TCGA_clin$histological_stage)
```

### Ref2
```{r}
table(Ref2_clin$histological_grade)
```

### Ref4
```{r}
table(Ref4_clin$histological_grade)
```

### RefGSE20271
```{r}
table(RefGSE20271_clin$histological_grade)
```

### RefGSE575678
```{r}
table(RefGSE575678_clin$histological_grade)
```


## RACE

### TCGA
```{r}
table(TCGA_clin$race)
```

### Ref2
```{r}
table(Ref2_clin$race)
```

### Ref4
```{r}
table(Ref4_clin$race)
```

### RefGSE20271
```{r}
table(RefGSE20271_clin$race)
```

### RefGSE575678
```{r}
table(RefGSE575678_clin$race)
```


## TREATMENT

### TCGA
```{r}
table(TCGA_clin$treatment)
```

### Ref2
```{r}
table(Ref2_clin$treatment)
```

### Ref4
```{r}
table(Ref4_clin$treatment)
```

### RefGSE20271
```{r}
sum(table(RefGSE20271_clin$treatment))
```

### RefGSE575678
```{r}
table(RefGSE575678_clin$treatment)
```


## TUMOR SIZE

### TCGA
```{r}
table(TCGA_clin$tumor_size)
```

### Ref2
```{r}
table(Ref2_clin$tumor_size_mm)
```

### Ref4
```{r}
table(Ref4_clin$tumor_size)
```

### RefGSE20271
```{r}
sum(table(RefGSE20271_clin$tumor_size))
```

### RefGSE575678
```{r}
table(RefGSE575678_clin$tumor_size_mm)
```

# COMBAT merged PCA
``````{r}
# TODO test this for gene intersection

# NOTES:
# > The databases not used for the resulting database could serve as validation
# > si se duplican muestras, quedarnos con 
# > Revisar los datos clínicos:
#    * Lo que diferencía a los 

# Revisar si los genes diferenciados son relevantes solamente para TNBC o para el cáncer en general o para el TNBC (pruebas estadísticas)
#   - Cluster 1 vs Base de datos original N (TNBC y otros)
#   - Plot: exprs Cluster A, Cluster B y otras DB originales

# TODO: DB before combat subir a Drive{r}
# TODO test this for gene intersection

#Intersect datasets and normalization
common_genes <- sort(intersect(intersect(intersect(rownames(Ref2_db), rownames(TCGA_db)), intersect(rownames(Ref4_db), rownames(RefGSE20271_db))),rownames(RefGSE575678_db)))

cy_mat <- cbind(TCGA_db[common_genes,], Ref2_db[common_genes,], Ref4_db[common_genes,], RefGSE20271_db[common_genes,], RefGSE575678_db[common_genes,])
cy_mat <- cy_mat[complete.cases(cy_mat),]

cy_mat_genes <- row.names(cy_mat)
cy_mat_samples <- colnames(cy_mat)
cy_mat <- normalize.quantiles(as.matrix(cy_mat))

cy_mat <- as.data.frame(cy_mat, row.names = cy_mat_genes)
colnames(cy_mat) <- cy_mat_samples

cy_mat = apply(cy_mat, 2, function(x) as.numeric(as.character(x)))

#MAD to filter genes and PCA
cy_mat_mad <- apply(cy_mat, 1, mad)
cy_mat_mad[is.na(cy_mat_mad)] <- 0

cy_mat_batch <- c(rep(1, ncol(TCGA_db)), rep(2, ncol(Ref2_db)), rep(3, ncol(Ref4_db)), rep(4, ncol(RefGSE20271_db)), rep(5, ncol(RefGSE575678_db)))
cy_mat_pca <- prcomp(cy_mat[which(cy_mat_mad > 1),])

#Combat
modcombat <- model.matrix(~1, data=data.frame(cmb=cy_mat_batch))
cy_mat_combat <- ComBat(dat=cy_mat, batch=cy_mat_batch, mod=modcombat, par.prior=TRUE, prior.plots=FALSE)
rownames(cy_mat_combat) <- cy_mat_genes

#MAD to filter genes and PCA
cy_mat_combat_mad <- apply(cy_mat_combat, 1, mad)
cy_mat_combat_pca <- prcomp(t(cy_mat_combat[which(cy_mat_combat_mad > 1),]))

#Plot PCA before and after Combat
setwd("C:/Users/victo/OneDrive/Documentos/TNBC_research/Database exploring")

#pdf("PCA_Before_Combat_3.pdf", width=9.2)
par(mfrow=c(1,2))
plot(cy_mat_pca$rotation[,1], cy_mat_pca$rotation[,2], xlab="PCA component 1", ylab="PCA component 2", main="Before Combat", col=cy_mat_batch, pch=19)
legend(x=0.06, y=0.2, legend=c("TCGA", "Ref 2", "Ref 4","RefGSE20271_db","RefGSE575678_db"), col=c(1,2,3,4,5), pch=19)
plot(cy_mat_combat_pca$rotation[,1], cy_mat_combat_pca$rotation[,2], xlab="PCA component 1", ylab="PCA component 2", main="Combat", col=cy_mat_batch, pch=19)
legend(x=-0.006, y=0.125, legend=c("TCGA", "Ref 2", "Ref 4","RefGSE20271_db","RefGSE575678_db"), col=c(1,2,3,4,5), pch=19)
#dev.off()

```

# Joining clinical datasets

### TCGA
```{r}
head(TCGA_clin)
```

### Ref2
```{r}
head(Ref2_clin)
```

### Ref4
```{r}
head(Ref4_clin)
```

### RefGSE20271
```{r}
head(RefGSE20271_clin)
```

### RefGSE575678
```{r}
head(RefGSE575678_clin)
```


## merging

First, we need to clean the datasets to ensure matching
```{r}
# Renaming a column
colnames(TCGA_clin)[12] <- "histological_grade"
colnames(TCGA_clin)
```


```{r}
class(TCGA_clin$histological_grade) = "character"
class(Ref2_clin$histological_grade) = "character"
class(Ref4_clin$histological_grade) = "character"
class(RefGSE20271_clin$histological_grade) = "character"
class(RefGSE575678_clin$histological_grade) = "character"

class(TCGA_clin$age) = "character"
class(Ref2_clin$age) = "character"
class(Ref4_clin$age) = "character"
class(RefGSE20271_clin$age) = "character"
class(RefGSE575678_clin$age) = "character"

class(Ref2_clin$tumor_stage) = "character"
class(Ref4_clin$tumor_stage) = "character"
```

Merging the datasets
```{r}

2
library(dplyr)
combo = bind_rows(TCGA_clin, Ref2_clin, Ref4_clin, RefGSE20271_clin, RefGSE575678_clin)
combo <- filter(combo, rownames(combo) %in% cy_mat_samples)
head(combo)
```

# Plotting merged PCA

```{r}
library(plyr)

combo$histological_grade <- mapvalues(combo$histological_grade, c("stage i", "stage ia", "stage iv", "stage iia", "stage iib", "II INTERMEDIATE GRADE (MODERATELY DIFFERENTIATED)", "stage ii", "stage iiic", "III HIGH GRADE (POORLY DIFFERENTIATED)", "stage iiia", "stage iiib", "yes", "no", "N/A"), c("1", "1", "1", "2", "2", "2", "2", "3", "3", "3", "3", "Unknown", "Unknown", "Unknown"))

unique(combo$histological_grade)
```

```{r}
library(plyr)

combo$histological_type <-
  mapvalues(
    combo$histological_type,
    c(
      "infiltrating ductal carcinoma",
      "medullary carcinoma",
      "Breast Invasive Lobular Carcinoma",
      "Breast Mixed Ductal and Lobular Carcinoma",
      "Infiltrating Lobular Carcinoma",
      "IDC/ILC",
      "metaplastic carcinoma",
      "Breast Invasive Ductal Carcinoma",
      "Breast Invasive Mixed Mucinous Carcinoma",
      "mixed histology (please specify)",
      "Infiltrating Ductal Carcinoma",
      "IDC",
      "infiltrating lobular carcinoma",
      "Breast Invasive Cancer, NOS",
      "Breast",
      "other, specify",
      "Mixed",
      "Other",
      "Medullary",
      "Ductal/NST"
    ),
    c(
      "Ductal",
      "Medular",
      "Lobular",
      "Ductal and Lobular",
      "Lobular",
      "Ductal and Lobular",
      "Metaplastic",
      "Ductal",
      "Mixed (not specified)",
      "Mixed (not specified)",
      "Ductal",
      "Ductal",
      "Lobular",
      "Not specified",
      "Not specified",
      "Not specified",
      "Mixed (not specified)",
      "Not specified",
      "Medular",
      "Ductal"
    )
  )

unique(combo$histological_type)
```

```{r}
library(ggrepel)
library(factoextra)

## MAKE SURE TO TRANSPOSE THE EXPRESSION MATRIX
# TODO: PCA with the 1500 - 2000 most variable (std or MAD (median absolute deviation)), más diferentes
#eset_TNBC_mad_ordered <- keep_most_different_genes(eset_TNBC, method="mad")[1:2000,]

#MAD to filter genes and PCA
cy_mat_combat_mad <- apply(cy_mat_combat, 1, mad)
cy_mat_combat_pca <- prcomp(t(cy_mat_combat[which(cy_mat_combat_mad > 1),]))

combo_combat <- filter(combo, rownames(combo) %in% colnames(cy_mat_combat[which(cy_mat_combat_mad > 1),]))
histological_type <- as.factor(combo_combat$histological_type)
histological_grade <- as.factor(combo_combat$histological_grade)
age <- as.factor(combo_combat$age)
tumor_size <- as.factor(combo_combat$tumor_size_mm)
source_db <- as.factor(combo_combat$source_db)
#treatment <- as.factor(sampleInfo$treatment)
#methastasis <- as.factor(sampleInfo$methastasis)

## Join the PCs to the sample information
fviz_pca_ind(cy_mat_combat_pca, repel=T, col.ind=histological_type, label="none")
fviz_pca_ind(cy_mat_combat_pca, repel=T, col.ind=histological_grade, label="none")
fviz_pca_ind(cy_mat_combat_pca, repel=T, col.ind=age, label="none")
fviz_pca_ind(cy_mat_combat_pca, repel=T, col.ind=tumor_size, label="none")
fviz_pca_ind(cy_mat_combat_pca, repel=T, col.ind=source_db, label="none")
#fviz_pca_ind(pca, repel=T, col.ind=treatment, label="none")
#fviz_pca_ind(pca, repel=T, col.ind=methastasis, label="none")
```

```{r}
library(readr)
# store rownames in new column
combo_combat<- as.data.frame(combo_combat)
combo_combat$row.names <- row.names(combo_combat)

cy_mat_combat<- as.data.frame(cy_mat_combat)
cy_mat_combat$row.names <- row.names(cy_mat_combat)

cy_mat_combat_mad<- as.data.frame(cy_mat_combat_mad)
cy_mat_combat_mad$row.names <- row.names(cy_mat_combat_mad)

combo<- as.data.frame(combo)
combo$row.names <- row.names(combo)

setwd("C:\\Users\\victo\\OneDrive\\Documentos\\TNBC_research\\Clinical_data_comparison")
write_csv(x = combo_combat, file="clinical_data_combat_mad.csv")
write_csv(x = combo, file="clinical_data_combat.csv")
write_csv(x = cy_mat_combat, file="exprs_data_combat.csv")
write_csv(x = cy_mat_combat_mad, file="exprs_data_combat_mad.csv")
```

