---
title: "combat"
author: "Victoria Rios"
date: "3/22/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## ComBat
```{r}
# Import the libraries
options(stringsAsFactors=F)
library(limma)
library(sva)
library(preprocessCore)
```

```{r}
# Import the datasets
TCGA_db <-
  read.csv("~/TNBC_research/data/db_files/TCGA/TCGA_TNBC_exprs.csv",
           row.names = "row.names")
Ref2_db <-
  read.csv(
    "~/TNBC_research/data/db_files/Ref 2/Ref2_untar/brca_metabric/Ref2_TNBC_exprs.csv",
    row.names = "row.names"
  )
Ref3_db <-
  read.csv("~/TNBC_research/data/db_files/Ref 3/Ref3_untar/Ref3_TNBC_exprs.csv",
           row.names = "row.names")
Ref4_db <-
  read.csv(
    "~/TNBC_research/data/db_files/Ref 4/Ref4_untar/brca_mbcproject_wagle_2017/Ref4_TNBC_exprs.csv",
    row.names = "row.names"
  )
RefGSE20271_db <-
  read.csv("~/TNBC_research/data/db_files/GSE20271/GSE20271_TNBC_exprs.csv",
           row.names = "row.names")
RefGSE575678_db <-
  read.csv("~/TNBC_research/data/db_files/GSE575678/GSE75678_TNBC_exprs.csv",
           row.names = "row.names")
```

## Finding intersections with UpSetR

```{r}
# First, we create a dataframe per dataset, containing the unique genes of each one with a vector of 1s, to account for their frequency

TCGA_vec <-
  data.frame("TCGA_vec" = rep(1, length(row.names(TCGA_db))))
TCGA_vec$id <- row.names(TCGA_db)

Ref2_vec <-
  data.frame("Ref2_vec" = rep(1, length(row.names(Ref2_db))))
Ref2_vec$id <- row.names(Ref2_db)

Ref3_vec <-
  data.frame("Ref3_vec" = rep(1, length(row.names(Ref3_db))))
Ref3_vec$id <- row.names(Ref3_db)

Ref4_vec <-
  data.frame("Ref4_vec" = rep(1, length(row.names(Ref4_db))))
Ref4_vec$id <- row.names(Ref4_db)

RefGSE20271_vec <-
  data.frame("RefGSE20271_vec" = rep(1, length(row.names(RefGSE20271_db))))
RefGSE20271_vec$id <- row.names(RefGSE20271_db)

RefGSE575678_vec <-
  data.frame("RefGSE575678_vec" = rep(1, length(row.names(RefGSE575678_db))))
RefGSE575678_vec$id <- row.names(RefGSE575678_db)
```

```{r}
library(tidyverse)
#put all data frames into list
df_list <- list(TCGA_vec, Ref2_vec, Ref3_vec, Ref4_vec, RefGSE20271_vec, RefGSE575678_vec)      

#merge all data frames together
df_all = df_list %>% reduce(full_join, by='id')

#fill NAs with 0
df_all = replace(df_all, is.na(df_all), 0)

#drop the id column and set is as the rownames
row.names(df_all) <- df_all$id
cols_df_all <- colnames(df_all)
rows_df_all <- rownames(df_all)

df_all <- t(df_all)
df_all <- as.data.frame(df_all)
colnames(df_all) <- rows_df_all
rownames(df_all) <- cols_df_all

## Uncomment to save to a file
#write_csv(x = df_all, file="ComBat/genes_present_all.csv")

```

```{r}
# Change the class of the dataframe values to numeric
df_all <- as.data.frame(sapply(df_all[-2, ], as.numeric))
rownames(df_all) <- cols_df_all[-2]
```

```{r}
library(pheatmap)

df_all_mat <- as.matrix(df_all)
df_all_mat <- df_all_mat[, colSums(df_all_mat) > 2, drop = FALSE]

pheatmap(
  df_all_mat,
  main = "Gene intersection between TNBC datasets",
  annotation_names_col = F,
  show_colnames = F
)
```

## ComBat
``````{r}
#Intersect datasets and normalization
common_genes <-
  sort(intersect(intersect(
    intersect(rownames(Ref2_db), rownames(TCGA_db)),
    intersect(rownames(Ref4_db), rownames(RefGSE20271_db))
  ), rownames(RefGSE575678_db)))

cy_mat <-
  cbind(TCGA_db[common_genes, ],
        Ref2_db[common_genes, ],
        Ref4_db[common_genes, ],
        RefGSE20271_db[common_genes, ],
        RefGSE575678_db[common_genes, ])
cy_mat <- cy_mat[complete.cases(cy_mat), ]

cy_mat_genes <- row.names(cy_mat)
cy_mat_samples <- colnames(cy_mat)
#cy_mat <- normalize.quantiles(as.matrix(cy_mat))

cy_mat <- as.data.frame(cy_mat, row.names = cy_mat_genes)
colnames(cy_mat) <- cy_mat_samples

cy_mat = apply(cy_mat, 2, function(x)
  as.numeric(as.character(x)))

#MAD to filter genes and PCA
cy_mat_mad <- apply(cy_mat, 1, mad)
cy_mat_mad[is.na(cy_mat_mad)] <- 0

cy_mat_batch <-
  c(rep(1, ncol(TCGA_db)),
    rep(2, ncol(Ref2_db)),
    rep(3, ncol(Ref4_db)),
    rep(4, ncol(RefGSE20271_db)),
    rep(5, ncol(RefGSE575678_db)))
cy_mat_pca <- prcomp(cy_mat[which(cy_mat_mad > 1), ])

#Combat
modcombat <- model.matrix( ~ 1, data = data.frame(cmb = cy_mat_batch))
cy_mat_combat <-
  sva::ComBat(
    dat = cy_mat,
    batch = cy_mat_batch,
    mod = modcombat,
    par.prior = TRUE,
    prior.plots = FALSE
  )
rownames(cy_mat_combat) <- cy_mat_genes

#MAD to filter genes and PCA
cy_mat_combat_mad <- apply(cy_mat_combat, 1, mad)
cy_mat_combat_pca <-
  prcomp(t(cy_mat_combat[which(cy_mat_combat_mad > 1), ]))

cy_mat_combat_mad <- cy_mat_combat[which(cy_mat_combat_mad > 1), ]

#Plot PCA before and after Combat

## Uncomment to save to a file
pdf("PCA_Before_Combat_1.pdf", width = 9.2)

par(mfrow = c(1, 2))
plot(
  cy_mat_pca$rotation[, 1],
  cy_mat_pca$rotation[, 2],
  xlab = "PCA component 1",
  ylab = "PCA component 2",
  main = "Before Combat",
  col = cy_mat_batch,
  pch = 19
)
legend(
  x = 0.06,
  y = 0.2,
  legend = c("TCGA", "Ref 2", "Ref 4", "RefGSE20271_db", "RefGSE575678_db"),
  col = c(1, 2, 3, 4, 5),
  pch = 19
)
plot(
  cy_mat_combat_pca$rotation[, 1],
  cy_mat_combat_pca$rotation[, 2],
  xlab = "PCA component 1",
  ylab = "PCA component 2",
  main = "Combat",
  col = cy_mat_batch,
  pch = 19
)
legend(
  x = -0.006,
  y = 0.125,
  legend = c("TCGA", "Ref 2", "Ref 4", "RefGSE20271_db", "RefGSE575678_db"),
  col = c(1, 2, 3, 4, 5),
  pch = 19
)
dev.off()

```

Exporting the merged dataset
```{r}
library(readr)

# store rownames in new column
cy_mat_combat<- as.data.frame(cy_mat_combat)
cy_mat_combat$row.names <- row.names(cy_mat_combat)

cy_mat_combat_mad<- as.data.frame(cy_mat_combat_mad)
cy_mat_combat_mad$row.names <- row.names(cy_mat_combat_mad)

## Uncomment to save to a file
#write_csv(x = cy_mat_combat, file="ComBat/ComBat.csv")
#write_csv(x = cy_mat_combat_mad, file="ComBat/ComBat_mad.csv")
```

# Using TDM library
```{r}
library(TDM)

# The file with the reference data has samples represented as rows,
# and genes as columns, so this creates a column with gene names.
temp_cy_mat <- t(cy_mat)
ref_values = data.table::data.table(cbind(colnames(temp_cy_mat), temp_cy_mat))

# Set the name of the column with the genes.
data.table::setnames(ref_values, colnames(ref_values)[1], "gene")

## Uncomment to save to a file
#TDM::tdm_transform(ref_values, ref_file = "ComBat/cy_mat_db.csv")
```


```{r}
#Tissue Type
coadread_sample <- substr(colnames(coadread_rnaseq), 14, 16)
cy_mat_type <- rep("Cancer", ncol(cy_mat_combat))
cy_mat_type[c(grep("normal", geoGSE25071_clin$Tissue), (which(coadread_sample == "11A") +
                                                          120))] <- "Normal"

#Plot PCA Tissue Type
par(mfrow = c(1, 2))
plot(
  cy_mat_combat_pca$rotation[, 1],
  cy_mat_combat_pca$rotation[, 2],
  xlab = "PCA component 1",
  ylab = "PCA component 2",
  main = "Combat",
  col = cy_mat_batch,
  pch = 19
)
legend(
  x = -0.036,
  y = 0.135,
  legend = c("GSE25071", "GSE39084", "TCGA"),
  col = c(1, 2, 3),
  pch = 19
)
plot(
  cy_mat_combat_pca$rotation[, 1],
  cy_mat_combat_pca$rotation[, 2],
  xlab = "PCA component 1",
  ylab = "PCA component 2",
  main = "Combat",
  col = ifelse(cy_mat_type == "Normal", 1, 2),
  pch = 19
)
legend(
  "topright",
  legend = c("Normal", "Cancer"),
  col = 1:2,
  pch = 19
)
```

